Report on the Third Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE3)

### 2015-09-30 WSSSPE3/report/WSSSPE3_review.tex
starting WSSSPE3 report

\title[]{Report on the Second Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE2)}

\author{Daniel S. Katz$^{(1)}$, Sou-Cheng T. Choi$^{(2)}$, Nancy Wilkins-Diehr$^{(3)}$, Neil Chue Hong$^{(4)}$,
\\Colin C. Venters$^{(5)}$, James Howison$^{(6)}$, Frank Seinstra$^{(7)}$, Matthew Jones$^{(8)}$,
\\Karen Cranston$^{(9)}$, Thomas L. Clune$^{(10)}$, Miguel de Val-Borro$^{(11)}$, Richard Littauer$^{(12)}$}
%
\thanks{{}$^{(1)}$ Computation Institute, 
University of Chicago \& Argonne National Laboratory, Chicago, IL, USA; \url{dsk@uchicago.edu}}
%
\thanks{{}$^{(2)}$ NORC at the University of Chicago and Illinois Institute of Technology, Chicago, IL, USA; \url{sctchoi@uchicago.edu}}
%
\thanks{{}$^{(3)}$ University of California-San Diego, San Diego, CA, USA; \url{wilkinsn@sdsc.edu}}
%
\thanks{{}$^{(4)}$ Software Sustainability Institute, 
University of Edinburgh, Edinburgh, UK; \url{N.ChueHong@software.ac.uk}}
%
\thanks{{}$^{(5)}$ University of Huddersfield, School of Computing and Engineering, Huddersfield, UK; \url{C.Venters@hud.ac.uk}}
%
\thanks{{}$^{(6)}$ University of Texas at Austin, Austin, TX, USA; \url{jhowison@ischool.utexas.edu}}
%
\thanks{{}$^{(7)}$ Netherlands eScience Centre, Amsterdam, Netherlands; \url{F.Seinstra@esciencecenter.nl}}
%
\thanks{{}$^{(8)}$ National Center for Ecological Analysis and Synthesis, Santa Barbara, CA, USA; \url{jones@nceas.ucsb.edu}}
%
\thanks{{}$^{(9)}$ National Evolutionary Synthesis Center, Durham, NC, USA; \url{karen.cranston@nescent.org}}
%
\thanks{{}$^{(10)}$ NASA Goddard Space Flight Center, Greenbelt, MD, USA; \url{Thomas.L.Clune@nasa.gov}}
%
\thanks{{}$^{(11)}$ Department of Astrophysical Sciences, 
Princeton University, Princeton, NJ, USA; \url{mdevalbo@astro.princeton.edu}}
%
\thanks{{}$^{(12)}$ University of Saarland, Germany; \url{richard.littauer@gmail.com}}
%

\begin{abstract}
This technical report records and discusses the Second Workshop on Sustainable
Software for Science: Practice and Experiences (WSSSPE2). 
%The workshop used an
%alternative submission and peer-review process, which led to a set of papers
%divided across five topic areas: 
The report includes a description of the alternative, experimental submission
and review process, two workshop keynote presentations, a series of lightning
talks, a discussion on sustainability, and five discussions from the topic areas
of exploring sustainability; software development experiences; credit \&
incentives; reproducibility \& reuse \& sharing; and code testing \& code
review. For each topic, the report includes a list of tangible actions that were
proposed and that would lead to potential change.
%
The workshop recognized that reliance on scientific software
is pervasive in all areas of world-leading research today. The workshop
participants then proceeded to explore different perspectives on the concept of
sustainability. Key enablers and barriers of sustainable scientific software
were identified from their experiences. In addition,
recommendations with new requirements such as software credit files and software
prize frameworks were outlined for improving practices in sustainable software
engineering.
%
There was also broad consensus that formal
training in software development or engineering was rare among the
practitioners. Significant strides need to be made in building a sense of
community via training in software and technical practices, on increasing their
size and scope, and on better integrating them directly into graduate education
programs.
%
Finally, journals can define and publish policies to improve reproducibility, whereas
reviewers can insist that authors provide sufficient information and access to
data and software to allow them reproduce the results in the paper. Hence a list of
criteria is compiled for  journals to provide to reviewers so as to make it easier to
review software submitted for publication as a ``Software Paper.''

@@

\maketitle
\newpage

@@

The Second Workshop on Sustainable Software for Science: Practice and
Experiences
(WSSSPE2)\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe2/}} was
held on 16 November, 2014 in the city of New Orleans, Louisiana, USA, in
conjunction with the International Conference for High Performance Computing,
Networking, Storage and Analysis
(SC14)\footnote{\url{http://sc14.supercomputing.org}}. WSSSPE2 followed the
model of a general initial workshop,
WSSSPE1\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe1/}}~\cite{WSSSPE1-pre-report,WSSSPE1},
which co-occurred with SC13, and a focused workshop,
WSSSPE1.1\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe1-1/}},
which was organized in July 2014 jointly with the SciPy
conference\footnote{\url{https://conference.scipy.org/scipy2014/participate/wssspe/}}.

@@

WSSSPE2 began with a call for papers~\cite{WSSSPE2-pre-report}. Based on the
goal of encouraging a wide range of submissions from those involved in software
practice, ranging from initial thoughts and partial studies to mature
deployments, but focusing on papers that are intended to lead to changes, the
organizers wanted to make submission as easy as possible. The call for papers
stated:

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Keynotes} \label{sec:keynotes}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{comment}
\note{Lead: Choi. \href{http://tinyurl.com/q6ew45v}{Google Notes}.
%https://docs.google.com/document/d/1-BxkYWDQ6nNNBXBStUL0xcKF9qCTlEALwf928J_MemI/edit?usp=sharing
\href{http://tinyurl.com/mnenzms}{Abstracts and slides}}
\end{comment}

The workshop featured two keynote addresses. In the opening keynote
presentation, Kaitlin Thaney of the Mozilla Science Lab talked about her
organization's work and policy to enable and support sustainable and
reproducible scientific research through the open web. The second keynote
speaker was Neil Chue Hong of Software Sustainability Institute. He shed
light on how scientific software is prevalently driving advances in many science
and engineering fields. Both keynote speeches spawned further discussion among
workshop participants on the crucial notion of \emph{software sustainability}
in the theme of our workshop.

%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
\subsection{Kaitlin Thaney, Designing for Truth, Scale, and Sustainability~\cite{Thaney_slides}}
\label{keynote1}
%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

@@

\begin{comment}
(Kaitlin didn't really say why? Were the tools used? Did they immediately
bit-rot? Was their development unsustained?)-should ask question about this. Is
there a caBIG report available? Quotes from Andrea Califano/Joe Gray, looks like
there is: \url{http://tinyurl.com/qdodbo7}.
%http://www.informationweek.com/architecture/report-blasts-problem-plagued-cancer-research-grid/d/d-id/1097068

What do we mean by open research? In regard to community/technology/practices

Inefficiency cartoon:  \url{http://www.xkcd.com/1445/}.
\end{comment}

@@

%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
\subsection{Neil Chue Hong, We are the 92$\%$~\cite{Hong_slides}}
\label{keynote2}
%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
\begin{comment}
In a recent survey of UK research-intensive universities, 92\% of researchers
said they used research software and 68\% said their research would be impossible
without software. Yet 71\% have had no formal software training, and few are
ready to apply many of the things we take for granted such as testing or
virtualization. WSSSPE represents the pinnacle of what we understand to be the
best practice around scientific software in our community. My talk will
challenge the workshop participants to come up with ways of taking this best
practice to those 92\% of researchers in a way that will lead to maximum benefit
to the scientific community.
 \end{comment}

%\url{http://dx.doi.org/10.6084/m9.figshare.1243288}

Neil Chue Hong is the Director of the Software Sustainability Institute (SSI) in the
United Kingdom (UK). The SSI was founded to support the UK's research software
community by cultivating better, more sustainable research software to enable
world-class research. Chue Hong's keynote started by making the point
that the use of -- and reliance on -- software is pervasive in all areas of
world-leading research, showing examples from disciplines as diverse as
humanities and high-energy physics, quoting Kerstin Kleese Van Dam of the
Pacific Northwestern National Laboratory via a petition
campaign\footnote{\url{http://tinyurl.com/nkn5tnv}}
%https://www.change.org/p/everyone-in-the-research-community-we-must-accept-that-software-is-fundamental-to-research-or-we-will-lose-our-ability-to-make-groundbreaking-discoveries}}
at change.org, ``\emph{Today there are very few science areas left which do not
rely on IT and thus software for the majority of their research work. More
importantly key scientific advances in experimental and observational science
would have been impossible without better software.}'' He also cited Daniel
Katz, Software Infrastructure for Sustained Innovation (SI2) Program Director of
the NSF, ``\emph{Scientific discovery and innovation are advancing along
fundamentally new pathways opened by development of increasingly sophisticated
software. Software is an integral enabler of computation, experiment and theory,
and directly responsible for increased scientific productivity and enhancement
of researchers' capabilities.}''

Chue Hong drew attention to the issue that in the cyber-infrastructure and
high-performance community, hundreds of thousands
of researchers developing software are all too often disregarded or considered the long tail.
Actually, the numbers point to the fact that they are the mainstream. %and we
%(referring to people including the audience of the talk) are the elite minority.
He emphasized that software is no longer special; it is both essential to and
common in scientific research. A 2014
survey\footnote{\url{http://tinyurl.com/ooajs7m}}
%http://www.software.ac.uk/blog/2014-12-04-its-impossible-conduct-research-without-software-say-7-out-10-uk-researchers}}
conducted by the SSI polled researchers from 15 research-intensive UK
universities (406 respondents covering a representative range of funders,
disciplines, and seniority). The survey reported that 92\% respondents confirmed the use of research
software and 89\% affirmed that it would be impossible or difficult to conduct
research without software. Nevertheless the British research community is just
starting to understand the magnitude of the issue. Whilst many researchers make
use of software such as MATLAB, SPSS, and Excel, data from the aforementioned
SSI survey shows that over half (56\%) of respondents developed their own
research software (which equates to over 140,000 researchers if extrapolated
across the UK) and yet 71\% of all UK researchers had no formal software-development training, having
to rely on their own coding skills.

Examining another aspect of the size of the research software community,
Chue Hong noted that the costs of software-reliant research in the UK
included \textsterling 840 million of investment in the financial year
2013--2014, and this amount has risen by 3\% on average over the past four
years. About 30\% of total research investment has been spent on research that
relies on software over the last four financial years. These numbers stemmed
from an analysis by the SSI of data from $49,650$ grant titles and abstracts published on
Gateway to Research between years 2010 and 2014. A similar analysis of university
jobs advertised in the same period discovered that despite this investment, only
4\% of positions were software development related, and of these only 17\% were
explicitly named as a software developer or software engineer positions: the vast
majority being advertised as research associate or research assistant positions.
This in turn leads to the issue of career paths for those bridging the research
and software worlds, who are essential to support the use and further
development of research software, a point highlighted by a graphic showing UK
STEM graduate career paths\footnote{Source: The Scientific Century, Royal
Society, 2010 (revised to reflect first stage clarification from ``What Do PhD's
Do?'' study)} showing that only 3.5\% were able to secure permanent positions.

To conclude, Chue Hong led the audience in discussing the following questions:
What are \emph{we} going to \emph{do} to help and benefit the 92\% of researchers
who rely on software? Who do we need to persuade? What are the incentives we
need to put in place? Finally, he challenged the workshop participants to change
the current deficient practices in research and academia.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Lightning Talks} \label{sec:lightning}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{comment}
\note{Lead: Choi. Volunteers welcome.
\href{http://wssspe.researchcomputing.org.uk/wssspe2/lightning-talks/}{Slides.}}

@@

\smallskip

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Defining Sustainability} \label{sec:defining}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%\todo{writing led by Dan - other contributions welcome}

In the first interactive session, the attendees divided themselves into groups
to discuss software sustainability. They were asked to
\begin{enumerate}
\item discuss what the term ``software sustainability'' meant to them

\item determine three things they considered to be significant enablers of
software sustainability

\item determine three things they considered to be significant barriers to
software sustainability.
\end{enumerate}
Once each group had come up with answers, all the answers were compiled, and the
attendees voted on which they thought were important by a show of hands.

The general responses to what software sustainability meant were:
\begin{itemize}
\item keeping software scientifically useful
\item separating techniques in code from knowledge in code
\item that an adequately large community finds value in software and is willing
to sustain it.
\end{itemize}
%\katznote{any discussion of this?}

The enablers of and barriers to software sustainability, roughly ranked by
attendee votes are shown in
{\tablename}s~\ref{tb:software_sustainability_enablers} and
\ref{tb:software_sustainability_barriers}, respectively.\footnote{A few other
items were suggested as barriers, but were not voted on due to lack of time in
the session:
layering up dependencies;
using software past its sustainable life;
using software past its usable life;
inertia for accepted answers versus wrong or right answers;
monolithic or poor code; and
need to restructure code when hardware/software/libraries change.
}
%\katznote{any discussion of these tables?}

\begin{table*}[ht]
\centering
\caption{Enablers of software sustainability, with 0 to 10 `*'s roughly
indicating the fraction of attendees who voted for an item as important.}
\label{tb:software_sustainability_enablers}
  \begin{small}
  \begin{tabular}{ | p{1.65cm} | p{12.0cm} |}
   %  \begin{tabular}{ | p{1.9cm} | p{11.5cm} |}
    \hline
  Importance & Item \\ \hline \hline
********** & healthy and vibrant communities; vibrant community to champion software \\ \hline
********** & designing for growth and extension -- open development \\ \hline
******* & culture in community for reuse \\ \hline
**** & portability \\ \hline
**** & culture in development community to support transition between developers \\ \hline
*** & interdisciplinary people: science + IT experience \\ \hline
** & planning for end of life \\ \hline
** & make smart choices about dependencies \\ \hline
* & thinking of software as product lines -- long term vs.~short term view \\ \hline
 & not all communities need new software \\ \hline
 & converting use into resources \\ \hline
    \end{tabular}
    \end{small}
\end{table*}

\begin{table}[ht]
\caption{Barriers to software sustainability, with 0 to 10 `*'s roughly
indicating the fraction of attendees who voted for an item as important.}
\label{tb:software_sustainability_barriers}
 \centering
  \begin{small}
  \begin{tabular}{ | p{1.65cm} | p{12.0cm} |}
% \begin{tabular}{ | p{1.9cm} | p{11.5cm} |}
   \hline
  Importance & Item \\ \hline \hline
******* & lack of incentives, including promotion and tenure process; promotion
and tenure process in academic is incompatible with sustainability \\ \hline
*****  & absent or poor documentation \\ \hline
***** & funding to ensure sustainability is difficult to obtain \\ \hline
**** & developers are not computer scientists; don't have software engineering
practices (in particular, those needed to scale-up projects to support and be
developed by a large sustainable community) \\ \hline
*** & overreliance on one or two people -- the `bus test'\footnote{``Bus test''
in Table~\ref{tb:software_sustainability_barriers} refers to the smallest number
of key people a project would need to lose to become non-viable (the larger the
number, the healthier the project).} \\ \hline
** & rate of change of underlying technologies \\ \hline
** & lack of business models for sustainability \\ \hline
* & lack of training for how to build sustainability into the system \\ \hline
 & maintenance for software is not visible, appears to ``just happen'' \\ \hline
 & licensing issues \\ \hline
 & staff turnover -- lack of continuity \\ \hline
\end{tabular}
 \end{small}
\end{table}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Exploring Sustainability} \label{sec:exploring}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\todo{Colin led writing of this section.}

%\note{\href{http://tinyurl.com/mpbhvyb}{Google doc notes}}
%https://docs.google.com/a/uchicago.edu/document/d/10XkshP3YjXnA5JQgP-UFHbOxshWczzehWGSo6V7bNH4/edit

Six papers were categorized under the theme of Exploring Sustainability. The
group included four of the authors from the six papers submitted and an
additional number of participants who expressed an interest in the theme at the
workshop. Each paper had a different perspective on the concept of
sustainability, which ranged from the sustainment of communities to defining
sustainability as a first-class, composite non-functional requirement.

%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
\subsection{Discussion and Actions}
%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
\subsubsection{Discussion}

Each author was invited to outline the key action from their paper as a
potential discussion point for the group; where the authors were not present,
the group facilitators outlined the actionable points from their papers. The key
actions from the six papers included recommendations for improving practice in
sustainable software engineering~\cite{wssspe2_rosada_de_souza, wssspe2_downs};
development of \emph{Software and Infrastructure as a Service} as a mechanism for
fostering sustainable science communities~\cite{wssspe2_patra}; developer
incentives for code contributions to open source projects~\cite{wssspe2_pierce};
establishment of a set of software engineering principles based on scalability,
reproducibility, and energy efficiency~\cite{wssspe2_shi}; and applying software
architectures as a mechanism for architectural-level reasoning about
sustainability~\cite{wssspe2_venters}. The group took the position of viewing
sustainability from the perspective of addressing the challenges related to the
development, deployment, and maintenance of reusable
software~\cite{WSSSPE1}.

The principal focus of the initial discussion
considered how to foster cultural change towards developing sustainable software
in academic environments. It was suggested that a new requirement driven by the
agencies funding research projects where software was an intrinsic part of
enabling the research program would be an additional element within the grant
proposal to provide a sustainability plan for sustaining the software. This type
of initiative would provide the necessary incentive and motivation for
researchers to consider how to sustain their software beyond the lifetime of the
project.

The discussion then focused on a need for a common language and a definition on
the concept of sustainability that moved beyond the current fuzzy definitions
where time was the simple measurement of sustainment. It was suggested that
there was a need to identify tangible actions that underpin sustainability that
developers could incorporate into the development stream of their software. This
prompted a debate regarding whether sustainability should be considered as a
non-functional requirement or software quality as defined within ISO/IEC
25010~\cite{iso25010}. The focus of the discussion was based on the paper by
Venters et al.~\cite{wssspe2_venters}, which suggested that sustainability is a
first-class, composite, non-functional requirement composed of a number of
sub-characteristics. It was generally agreed that maintainability and
extensibility were key qualities underlying sustainability. In addition, the
group also discussed what other non-functional requirements would contribute to
the development of sustainable software, e.g., reusability and scalability.
However, it was recognized that there was a need to identify appropriate metrics
and measures.

The group also discussed whether the concept of sustainability itself was a
barrier to achieving sustainable software. It was suggested that ``the first
rule of software sustainability is do not talk about software
sustainability.'' Instead there should be a focus on best software engineering
practice. Playing devil's advocate, it was asked that if the focus on the
concept of sustainability was ignored, what current software engineering
practices and principles could be utilized by software developers and domain
scientists to achieve sustainability? This raised the question of why existing
software engineering knowledge, such as that contained in
SWEBOK~\cite{Bourque2014}, was largely ignored and to what extent the
environment has a strong influence on practice within the scientific and
engineering community. As a result, there is a need to identify best practices
and reach a consensus beyond the theories that underpin the discipline of
software engineering. Similarly, how could we translate or distill some of the
key building blocks that underpin software engineering practice?

A final key point of the discussion was the role that software design and
patterns play before committing to writing a line of code. It was suggested that
modeling must play a major role in attaining sustainable software. The point was
raised that design involves making decisions and a mechanism would be necessary
for capturing these decision points. This introduced the idea of software
provenance that moved beyond commits in software repositories to how to capture
and maintain relationships between sources and design decisions to prevent
knowledge vaporization. Whether this could be achieved through software
architectures is an open-research challenge.

\subsubsection{Actions}

The main action to come from the group was a proposal to identify the ten best
software engineering practices similar to Philip Bourne's approach of Ten Simple
Rules~\cite{bourne-ten-simple-rules}.

%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
\subsection{Papers}
%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
The papers that were discussed in the Exploring Sustainability group are:
\begin{itemize}
\item Mario Rosado de Souza, Robert Haines, and Caroline Jay. Defining
sustainability through Developers' Eyes: Recommendations from an Interview
Study~\cite{wssspe2_rosada_de_souza}

\item Robert Downs, W. Christopher Lenhardt, Erin Robinson, Ethan Davis, and
Nicholas Weber. Community Recommendations for Sustainable Scientific
Software~\cite{wssspe2_downs}

\item Abani Patra, Matthew Jones, Steven Gallo, Kyle Marcus, and Tevfik Kosar.
Role of Online Platforms, Communications and Workflows in Developing Sustainable
Software for Science Communities~\cite{wssspe2_patra}

\item Marlon Pierce, Suresh Marru, and Chris Mattmann. {WSSSPE2}: Patching It Up,
Pulling It Forward~\cite{wssspe2_pierce}

\item Justin Shi. Seeking the principles of sustainable software
engineering~\cite{wssspe2_shi}

\item Colin C. Venters, Michael K. Griffiths, Violeta Holmes, Rupert R. Ward, and
David J. Cooke. The Nebuchadnezzar Effect: Dreaming of Sustainable Software
through Sustainable Software Architectures~\cite{wssspe2_venters}
\end{itemize}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Software Development Experiences} \label{sec:devel}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\todo{Nancy led writing of this section, with Subgroup A part from Matt Jones}

%\note{\href{http://tinyurl.com/pn4eq8z}{Google doc notes}}
%https://docs.google.com/a/uchicago.edu/document/d/1_q0jmiEPNFRGAEtGLk_QxQ1vVokgTDYruTGURowNGkY/edit

Of the short actionable papers that would lead to improvements for sustainable
software science, 11 submissions were categorized in the Software Development
Experiences group. Because of the large number, we split these into two
subgroups prior to the event. Some common themes helped in this division. For
example, several papers that addressed education and training issues, including
best practices and case studies, were grouped together. Others discussed
experiences with registries, developer collectives and specific examples of
successful, sustainable software (including a valuable industry perspective).

%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
\subsection{Discussion and Actions}
%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

Subgroup A consisted of ten participants who discussed papers surrounding training
and successful community software initiatives.
Subgroup B had six participants who discussed four papers. Because of the nature
of the papers, training emerged as a common theme. However the conversation was
wide-ranging, including incentives, reproducibility, and funding to promote
sustainability.  In the end, both groups discussed training, though from somewhat different
points of view and resulting in somewhat different suggested actions.

\subsubsection{Subgroup A Discussion}

%\note{\href{https://docs.google.com/document/d/1n413NcFPDyDGb27tEp-ujryiRhz-ZaAhmUapWRYmh1g/edit}{Google doc notes}}

Group discussion started with a position statement by each person surrounding
what they had learned from their software development experiences and how those
lessons might be translated into actionable outcomes. Participants came from a
range of backgrounds, and represented multiple software and training
initiatives, including Software Carpentry, Data Carpentry, Open Science for
Synthesis (OSS), the Community Surface Dynamics Modeling Systems group, ROpenSci,
DataONE, the HDF Group, and others. Some of these software experiences were
focused on development of new products for use in the sciences (e.g., ROpenSci,
CSDMS), and these recounted the difficulties of engaging with disciplinary
scientists in writing software. Software was clearly utilized broadly across the
various science disciplines represented, and it was developed within those
disciplines as well. Some researchers created software for statistical analysis
and modeling, while others used it to control instrumentation, collate data
across networks, collect information from respondents, and many other uses.
There was broad consensus that, within the disciplines represented, formal
training in any type of software development or engineering was rare among the
practitioners; most are self-taught, and develop software to get another job
done. Any ancillary utility of the software outside of the specific science
target was generally unplanned and few researchers would want to invest more
time to make their own software more re-usable.

%Despite agreement that this body of disciplinary software needed improvement,
There was general agreement that this body of disciplinary software improvement
needed to be understood in terms of the scientific productivity that could be
achieved. A software maturity model is needed for science software, but it needs
to be introduced in a way that fits the culture of science, which largely thinks
of software as a tool, rather than a product itself. The group was in general
consensus that more widespread training in software practice is needed within
the domain sciences, and several of the participants were involved in efforts
along these lines. Participants felt that projects that build a sense of
community via training in software and technical practices would have the most
success in changing practices in that community, but that there was a need for a
managed introduction of these practices. Participants also recognized that these
could not be one-time, one-off training opportunities, as software and
technologies change rapidly over time. For example, while today Software
Carpentry is focused on teaching version control via Git, there has been a rapid
evolution from RCS to CVS to SVN to Git over a short time frame, and thus
communities should expect the need to train for adaptability and a changing
technology tool chain, rather than assume that these technologies will stay
fixed. Thus, although short term training that introduces specific tools was
considered highly valuable, these trainings were also not considered sufficient
to engender the changes in software practice that were deemed necessary.
Combining the need for changes in practice to be introduced incrementally with
the need to minimize the divergence of training from direct science goals and
the difficulties of training for a rapidly evolving technology space,
participants concluded that multiple training efforts that targeted different
parts of the spectrum were needed. Short-term courses introducing immediately
useful skills needs to be offered alongside more in-depth courses on software
engineering and practice that allow students to adapt to a changing landscape.

Finally, after agreeing that these complementary training models were needed,
the group discussed sustainability of training, and how the leading groups in
this space are teaching only a small fraction of the community that needs and
wants training. Most graduate programs in the sciences do not currently
incorporate these approaches in their graduate curricula, although there is an
increasing number of quantitatively oriented courses around analysis and
modeling. These still, however, generally omit engineering practices such as
version control, unit and regression testing, and software modularity and
abstraction, often because the instructors themselves in the domain sciences are
not familiar with these techniques. Thus, students who are being trained in
these approaches are doing so through short 2--3 day training workshops such as
Software and Data Carpentry, rather than through semester-long graduate
education courses at their universities, which tend to focus on statistical and
modeling techniques. Hybrid programs like the three-week Open Science for
Synthesis (OSS) training that combine the three (science, quantitative
techniques, and software engineering) into an holistic course serve part of the
need but reach only a few researchers at this time. Thus, the group concluded
that significant strides needed to be made in coordinating these trainings so
that they are complementary, on increasing their size and scope, and on better
integrating them directly into graduate education programs.

>>>>

\title[]{Report on the Third Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE3)}

\author{tbd by writing and organizing}

%\author{Daniel S. Katz$^{(1)}$, Sou-Cheng T. Choi$^{(2)}$, Nancy Wilkins-Diehr$^{(3)}$, Neil Chue Hong$^{(4)}$,
%\\Colin C. Venters$^{(5)}$, James Howison$^{(6)}$, Frank Seinstra$^{(7)}$, Matthew Jones$^{(8)}$,
%\\Karen Cranston$^{(9)}$, Thomas L. Clune$^{(10)}$, Miguel de Val-Borro$^{(11)}$, Richard Littauer$^{(12)}$}
%%
%\thanks{{}$^{(1)}$ Computation Institute, 
%University of Chicago \& Argonne National Laboratory, Chicago, IL, USA; \url{dsk@uchicago.edu}}
%%
%\thanks{{}$^{(2)}$ NORC at the University of Chicago and Illinois Institute of Technology, Chicago, IL, USA; \url{sctchoi@uchicago.edu}}
%%
%\thanks{{}$^{(3)}$ University of California-San Diego, San Diego, CA, USA; \url{wilkinsn@sdsc.edu}}
%%
%\thanks{{}$^{(4)}$ Software Sustainability Institute, 
%University of Edinburgh, Edinburgh, UK; \url{N.ChueHong@software.ac.uk}}
%%
%\thanks{{}$^{(5)}$ University of Huddersfield, School of Computing and Engineering, Huddersfield, UK; \url{C.Venters@hud.ac.uk}}
%%
%\thanks{{}$^{(6)}$ University of Texas at Austin, Austin, TX, USA; \url{jhowison@ischool.utexas.edu}}
%%
%\thanks{{}$^{(7)}$ Netherlands eScience Centre, Amsterdam, Netherlands; \url{F.Seinstra@esciencecenter.nl}}
%%
%\thanks{{}$^{(8)}$ National Center for Ecological Analysis and Synthesis, Santa Barbara, CA, USA; \url{jones@nceas.ucsb.edu}}
%%
%\thanks{{}$^{(9)}$ National Evolutionary Synthesis Center, Durham, NC, USA; \url{karen.cranston@nescent.org}}
%%
%\thanks{{}$^{(10)}$ NASA Goddard Space Flight Center, Greenbelt, MD, USA; \url{Thomas.L.Clune@nasa.gov}}
%%
%\thanks{{}$^{(11)}$ Department of Astrophysical Sciences, 
%Princeton University, Princeton, NJ, USA; \url{mdevalbo@astro.princeton.edu}}
%%
%\thanks{{}$^{(12)}$ University of Saarland, Germany; \url{richard.littauer@gmail.com}}
%%

\begin{abstract}
\todo{need to update for WSSSPE3}
This technical report records and discusses the Second Workshop on Sustainable
Software for Science: Practice and Experiences (WSSSPE2). 
%The workshop used an
%alternative submission and peer-review process, which led to a set of papers
%divided across five topic areas: 
The report includes a description of the alternative, experimental submission
and review process, two workshop keynote presentations, a series of lightning
talks, a discussion on sustainability, and five discussions from the topic areas
of exploring sustainability; software development experiences; credit \&
incentives; reproducibility \& reuse \& sharing; and code testing \& code
review. For each topic, the report includes a list of tangible actions that were
proposed and that would lead to potential change.
%
The workshop recognized that reliance on scientific software
is pervasive in all areas of world-leading research today. The workshop
participants then proceeded to explore different perspectives on the concept of
sustainability. Key enablers and barriers of sustainable scientific software
were identified from their experiences. In addition,
recommendations with new requirements such as software credit files and software
prize frameworks were outlined for improving practices in sustainable software
engineering.
%
There was also broad consensus that formal
training in software development or engineering was rare among the
practitioners. Significant strides need to be made in building a sense of
community via training in software and technical practices, on increasing their
size and scope, and on better integrating them directly into graduate education
programs.
%
Finally, journals can define and publish policies to improve reproducibility, whereas
reviewers can insist that authors provide sufficient information and access to
data and software to allow them reproduce the results in the paper. Hence a list of
criteria is compiled for  journals to provide to reviewers so as to make it easier to
review software submitted for publication as a ``Software Paper.''

@@

\maketitle
%\newpage

@@

\todo{need to update for WSSSPE3}
The Second Workshop on Sustainable Software for Science: Practice and
Experiences
(WSSSPE2)\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe2/}} was
held on 16 November, 2014 in the city of New Orleans, Louisiana, USA, in
conjunction with the International Conference for High Performance Computing,
Networking, Storage and Analysis
(SC14)\footnote{\url{http://sc14.supercomputing.org}}. WSSSPE2 followed the
model of a general initial workshop,
WSSSPE1\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe1/}}~\cite{WSSSPE1-pre-report,WSSSPE1},
which co-occurred with SC13, and a focused workshop,
WSSSPE1.1\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe1-1/}},
which was organized in July 2014 jointly with the SciPy
conference\footnote{\url{https://conference.scipy.org/scipy2014/participate/wssspe/}}.

@@

\todo{need to update for WSSSPE3}
WSSSPE2 began with a call for papers~\cite{WSSSPE2-pre-report}. Based on the
goal of encouraging a wide range of submissions from those involved in software
practice, ranging from initial thoughts and partial studies to mature
deployments, but focusing on papers that are intended to lead to changes, the
organizers wanted to make submission as easy as possible. The call for papers
stated:

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Keynote} \label{sec:keynote}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\todo{need to update for WSSSPE3}
The workshop featured two keynote addresses. In the opening keynote
presentation, Kaitlin Thaney of the Mozilla Science Lab talked about her
organization's work and policy to enable and support sustainable and
reproducible scientific research through the open web. The second keynote
speaker was Neil Chue Hong of Software Sustainability Institute. He shed
light on how scientific software is prevalently driving advances in many science
and engineering fields. Both keynote speeches spawned further discussion among
workshop participants on the crucial notion of \emph{software sustainability}
in the theme of our workshop.

@@

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Lightning Talks} \label{sec:lightning}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\todo{need to update for WSSSPE3}
\begin{comment}
\note{Lead: Choi. Volunteers welcome.
\href{http://wssspe.researchcomputing.org.uk/wssspe2/lightning-talks/}{Slides.}}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Working Groups} \label{sec:WGs}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
